Palaeodistribution of crop progenitors and wild foods in the Southern Levant

### 2023-07-31 analysis/analysis.Rmd
Use rgbif instead of dismo

```{r dependencies}
library("dplyr")
library("ggplot2")
library("ggspatial")
library("parsnip")
library("purrr")
library("ranger")
library("rpaleoclim")
library("sf")
library("tibble")
library("tidyr")

@@

```{r data-occ}
# Download or read data from GBIF
# TODO: T. boeoticum ? (equate with T. urartu?)
gbif_path <- here::here("analysis", "data", "derived_data", "gbif")
occ <- bind_rows(
  fetch_gbif("Hordeum", "spontaneum*", ext = w_asia, gbif_path),
  #fetch_gbif("Bolboschoenus", "glaucus*", ext = w_asia, gbif_path),    
  #fetch_gbif("Bolboschoenus", "maritimus*", ext = w_asia, gbif_path),    
  #fetch_gbif("Scirpus", "maritimus*", ext = w_asia, gbif_path),
  #fetch_gbif("Triticum", "urartu*", ext = w_asia, gbif_path),
  #fetch_gbif("Triticum", "dicoccoides*", ext = w_asia, gbif_path),
  #fetch_gbif("Stipa", "*", ext = w_asia, gbif_path),
  #fetch_gbif("Zilla", "spinosa*", ext = w_asia, gbif_path),
  #fetch_gbif("Vitex", "*", ext = w_asia, gbif_path),
  #fetch_gbif("Fraxinus", "*", ext = w_asia, gbif_path)
)

# Code taxonomic groups for analysis
occ |>
  mutate(taxon = case_when(genus == "Hordeum" ~ "Hordeum spontaneum",
                           genus == "Bolboschoenus" ~ "Bolboschoenus sp.",
                           species == "Triticum urartu" ~ "Triticum urartu",
                           species == "Triticum dicoccoides" ~ "Triticum dicoccoides",
                           genus == "Stipa" ~ "Stipa sp.",
                           genus == "Zilla" ~ "Zilla spinosa",
                           genus == "Vitex" ~ "Vitex sp.",
                           genus == "Fraxinus" ~ "Fraxinus sp.")) ->
  occ

# TODO: convert to a test
if(any(is.na(occ$taxon))) stop("Taxon contains missing values!")

# Remove duplicate records within analytic groups
occ |>
  group_by(taxon) |>
  distinct(lon, lat, .keep_all = TRUE) |>
  ungroup() ->
  occ

# Convert to sf
occ |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) ->
  occ
```

```{r plot-occ-distribution}
ggplot() + 
  layer_spatial(ne_land, fill = "#ffffff") +
  annotation_spatial(s_levant, fill = alpha("red", 0.15), colour = "red") +
  layer_spatial(occ, alpha = 0.1, symbol = 16) +
  # layer_spatial(ne_countries, fill = NA) +
  coord_sf(xlim = w_asia[c("xmin", "xmax")], 
           ylim = w_asia[c("ymin", "ymax")],
           crs = utm37n,
           default_crs = latlong)
```

@@

occ |>
  mutate(present = TRUE) |>
  bind_rows(background_sample(s_levant_land, 
                              n_background,
                              coord_x = "lon",
                              coord_y = "lat", 
                              taxon = "Hordeum spontaneum", 
                              present = FALSE)) ->
  occ

@@

```{r data-paleoclim}
tibble(period = c("lgm", "hs1", "ba", "yds", "eh", "cur")) |>
  mutate(raster = map(period, paleoclim, "2_5m", 
                      region = buffer_bbox(w_asia, 10),
                      cache_path = here::here(".paleoclim_cache"))) -> # TODO: revert to default cache
  climate

@@

occ |>
  mutate(bio = terra::extract(training_climate, data.frame(lon, lat), 
                              method = "bilinear", ID = FALSE)) |>
  unpack(bio) |>
  ungroup() ->
  occ
```

>>>>

```{r dependencies}
library("dplyr")
library("ggplot2")
library("ggspatial")
library("here")
library("parsnip")
library("purrr")
library("ranger")
library("rpaleoclim")
library("sf")
library("tibble")
library("tidyr")

@@

```{r data-occ}
gbif_data <- function(scientificName, ...) {
  gbif_data <- rgbif::occ_data(scientificName = scientificName, ...)
  gbif_data$data
}

gbif_path <- here::here("analysis", "data", "derived_data", "gbif")

taxa <- c("Hordeum spontaneum" = "Hordeum spontaneum")
          # "Bolboshoenus spp." = "Bolboschoenus")

# Get occurrence data from GBIF
# Including only records with coordinates and no recorded issues with geospatial
#   data.
taxa |>
  map(gbif_data, hasCoordinate = TRUE, hasGeospatialIssue = FALSE, 
      geometry = w_asia, limit = 100000) |>
  list_rbind(names_to = "taxon") ->
  occ

# Exclude:
# * Coordinate uncertainty of 5 km or more (≈2.5 mins)
# * Fossil occurrences
occ |>
  filter(coordinateUncertaintyInMeters < 5000 | is.na(coordinateUncertaintyInMeters)) |>
  filter(basisOfRecord != c("FOSSIL_SPECIMEN")) ->
  occ

# Remove duplicated coordinates (as recommended by CITE)
occ |>
  group_by(taxon) |>
  distinct(decimalLongitude, decimalLatitude, .keep_all = TRUE) |>
  ungroup() ->
  occ

# Convert to sf
occ |>
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), 
           crs = 4326, remove = FALSE) ->
  occ
```

```{r plot-occ-distribution}
ggplot(occ) + 
  facet_wrap(vars(taxon)) +
  layer_spatial(ne_land, fill = "#ffffff") +
  annotation_spatial(s_levant, fill = alpha("red", 0.15), colour = "red") +
  layer_spatial(occ, alpha = 0.1, symbol = 16) +
  layer_spatial(ne_countries, fill = NA) +
  coord_sf(xlim = w_asia[c("xmin", "xmax")], 
           ylim = w_asia[c("ymin", "ymax")],
           crs = utm37n,
           default_crs = latlong)
```

@@

occ |>
  mutate(present = TRUE) |>
  bind_rows(background_sample(s_levant_land, 
                              n_background,
                              coord_x = "decimalLongitude",
                              coord_y = "decimalLatitude", 
                              taxon = "Hordeum spontaneum", 
                              present = FALSE)) ->
  occ

@@

```{r data-paleoclim}
tibble(period = c("lgm", "hs1", "ba", "yds", "eh", "cur")) |>
  mutate(raster = map(period, paleoclim, "2_5m", 
                      region = buffer_bbox(w_asia, 10),
                      cache_path = here(".paleoclim_cache"))) -> # TODO: revert to default cache
  climate

@@

occ |>
  mutate(bio = terra::extract(training_climate, data.frame(decimalLongitude, decimalLatitude), 
                              method = "bilinear", ID = FALSE)) |>
  unpack(bio) |>
  ungroup() ->
  occ
```
